### 2021-03-05 ch/2-related-work.tex
Use include instead of import
This has the advantage that I can compile one chapter at a time.

%
% =========================
%      old introduction
% =========================
% \section{Introduction}\label{introduction}
% Fairness in Machine Learning is a topic that has recently received increasing attention.
% Machine Learning has become very widespread in our society
% which brings some dangers with it.
% Decisions such as credit eligibility and hiring are increasingly done automatically.
% However, the algorithms that are used in these decisions can have strong biases
% that stem from the training data.
% Often, the algorithms are even more biased than the training data.
% \citet{kamishima2012fairness} give a great illustration for this
% (adapted from \citet{calders2010three}).
% The illustration is based on the Adult/Census Income dataset \citep{kohavi1996scaling}
% which contains records of individuals with thirteen attributes, among them age and gender.
% The task is to predict whether or not the individual earns more than \$50K per year.
% In the dataset, 32\% of the men have a positive label (\ie, they earn more than \$50K),
% but of the women only 11\% have a positive label.
% With this ratio of about 3:1, the dataset is already skewed.
% However, when training a \href{https://en.wikipedia.org/wiki/Naive_Bayes_classifier}{Na\"ive Bayes classifier} on the data,
% the ratio in the predictions is even more unfair:
% the amount of women who get a positive prediction is even lower than 11\%.
% If the Na\"ive Bayes classifier does not get access to the gender information,
% the predictions will be a bit fairer,
% but they are still more unfair than the original dataset labels.
%
% This phenomenon of predictions being more unfair than an already skewed dataset
% is due to the generalization that classifiers have to make.
% It is easier to classify all but the most outstanding women as low income
% than to learn a complicated decision boundary.
% Furthermore, removing the gender information does not solve the problem
% because the gender can be inferred from other data.
%
% The following is a review of the literature on Fairness in Machine Learning.
% Several different approaches to and definition of Fairness are presented.
% The focus is on identifying the major trends in the field from the beginning.
\section{\NoCaseChange{Two views of the dataset bias problem}}\label{two-views-of-the-dataset-bias-problem}
As hinted to in the introduction,
there is a divide in the relevant literature in how the problem of dataset bias is approached.
On the one hand, there is the ``ground-truth-centric'' view -- expressed through most of the introduction --
that the training set is biased but the deployment setting is not.
By \emph{deployment setting} I am here referring to the real distribution
that will be encountered by the machine learning system once it is in use.

@@

An instance of this setting has been formalized by \citet{blum2020recovering}.
In their formalization, the training set starts out unbiased,
but then, for samples with \(s=0\), class labels \(y\) are flipped from 1 to 0 with probability \(\nu\).
In addition, samples with \(s=0\) and \(y=0\) are dropped with probability \(1-\beta_\mathit{NEG}\)
and those with \(s=0\) and \(y=1\) are dropped with probability \(1-\beta_\mathit{POS}\).
Meanwhile, the test remains unbiased and is the basis for evaluating our models.

@@

\section{Precursors}\label{precursor}
The first published work to address the problem of biases in machine learning
was arguably the work by \citet{pedreshi2008discrimination}
with the main focus being data mining.
The first fundamental conclusion of the authors
is that it is not enough to take out the discriminatory features
(referred to as \emph{sensitive attributes} in other works) from the data.
That is, if the information about, for example, race
is removed from the data,
there is usually enough background information to reconstruct that attribute.
The experiments were done on the ``German credit dataset''
where the task is to classify people as a good or a bad credit risk.
In their context of data mining,
they distinguish between direct and indirect discrimination:
the former explicitly uses discriminatory features in the premise of the rule,
while the latter uses features that are closely associated with discriminatory features.
% The latter case can be
% detected by checking the other rules in the dataset.

@@

The authors then define a concrete measure of discrimination:
\begin{align}
  \label{eq:disc}
  Disc := P(y=1|s =1) - P(y=1|s = 0)
\end{align}
% (The paper uses a different notation.
% The notation here reflects the notation in later papers.)

This fairness metric corresponds to what is commonly referred to as \emph{demographic parity}.
It is zero when the individuals in both groups have the same chance to get a positive prediction.

@@

The second method in \citet{calders2010three}
is based on training two separate models for each sensitive attributes.
That is, the dataset is split in two
where one half contains all examples with \(s=0\), we call this dataset \(D_0\),
and the other half contains all examples with \(s=1\), called \(D_1\).
For prediction, we choose either the model that was trained on \(D_0\) or the one for \(D_1\),
depending on the sensitive attribute of the example.
This method relies on the availability of the sensitive attribute for prediction,
like the previous method.
The two models are separately tweaked to produce overall fair results.
In order to minimize the effect on the accuracy, the models are tweaked the same amount each,
but in opposite directions.
Conceptually, having two different Na\"ive Bayes model depending on the (binary) sensitive attribute
is equivalent to using one Na\"ive Bayes model
in which the sensitive attribute is connected to all other features.
This means that the bias is in all of the features and not just in the class label.
The latter was the assumption of the previously discussed model.
The difference between these assumptions is not explored in much detail in the paper.

The third methods introduces a new latent variable, called \(L\) in the paper.
\(L\) is intended to be an unbiased target class label
that replaces the biased class label \(y\) from the training data.
\(L\) can only be estimated and is unbiased
in the sense that it is statistically independent from sensitive attribute: \(P(L|s=0) = P(L|s=1)\).
The observed class label \(y\) then only depends on the latent label and the sensitive attribute.
\(L\) is found by using expectation maximization
to iteratively search for a value that maximizes the likelihood of the dataset.
The search is restricted by enforcing that \(L\) must be equal to \(y\) except
when \(s=0\) and \(y=0\) or when \(s=1\) and \(y=1\)
because these are the two cases where we expect discrimination.
There is other prior knowledge that can be incorporated into the search.
This method seems to have been the most sophisticated at that point in time.

@@

\begin{enumerate}
\item
  Prejudice:
  a statistical dependency between the sensitive attribute
  and the class label or the (non-sensitive) input features.
  If the dependency is with the class label, then the paper calls it \emph{indirect prejudice}.
  If it is with the non-sensitive features, they call it \emph{latent prejudice}.
  In line with \citet{pedreshi2008discrimination},
  \emph{direct prejudice} refers to classifiers that make direct use of the sensitive attribute.
  Indirect prejudice and latent prejudice can be quantified by
  the mutual information between \(y\) and \(s\) or \(x\) and \(s\) respectively.
  \citet{kamishima2012fairness} do not consider latent prejudice to be immediately harmful,
  but it can be a problem with respect to the protection of personal data and compliance with laws.
\item
  Underestimation:
  non-convergence of machine learning algorithm due to limited amount of data.
  This magnitude of this problem can be estimated
  by considering the difference between the actual training sample distribution
  and the distribution that the machine learning model has internalized.
  This is the underlying cause of models that make predictions
  that are even more unfair than the dataset.
\item
  Negative Legacy:
  sampling bias and wrong labels in the dataset.
  In contrast to the problem of \emph{prejudice},
  \emph{negative legacy} might not be detectable by analysing the dataset.
  \emph{Sampling bias}, meaning that certain data points simply are missing,
  and \emph{wrong labels} can only be corrected if other sources of information are available.
\end{enumerate}

We see that previous works nearly exclusively dealt with the first of these issues, \emph{prejudice},
and more precisely \emph{indirect prejudice}.
\citet{kamishima2012fairness} also focus on \emph{indirect prejudice};
essentially because it is the easiest to deal with, apart from \emph{direct prejudice}.
They use \acf{LR} as the basis for their fair classifier.
Their approach to enforcing fairness is qualitatively different from the previous proposals.
Instead of manipulating the dataset or using explicit algorithm to make a classifier fair,
\citet{kamishima2012fairness} treat the fairness constraint like a regulariser.
A term is added to the objective function
that estimates the mutual information between \(y\) and \(s\)
and gets minimized together with the other parts of the objective function.
This strategy leaves most of the work to the optimization that would need to be run anyway.
A factor in front of the regularization term determines how much to value fairness over accuracy.

@@

The proposed idea is then as follows: a classifier is fair
if and only if the predictive distributions for any two data points
are at least as similar as the two points themselves,
according to a given similarity measure for distributions
and a given similarity measure for data points.
They call this condition the \emph{Lipschitz condition}.
The authors propose two practical similarity measures for distributions
that are well-known in the respective literature.
In order to train a fair classifier,
the Lipschitz condition is then used as a constraint for the optimization.
This is similar to how \citet{kamishima2012fairness} used the fairness criterion as a regulariser.
However, \citet{dwork2012fairness} intend to realize it as a strict constraint.
Furthermore, they show that their criterion is compatible with Demographic Parity
but does not guarantee it in general.
When the two demographic groups are similar according to the individual similarity measure,
then the Lipschitz condition will lead to a good approximation of Demographic Parity.
If the demographic groups are very different,
then the Lipschitz criterion is essentially incompatible with Demographic Parity.

@@

\section{Fair representation}\label{sec:fair-representation}
\citet{zemel2013learning} expand on the idea of creating a \emph{fair representation}.
That is, the features of the data points are transformed such that all bias is removed
but still carry the information necessary for predicting the class label.
This is different from \citet{kamiran2009classifying} where the dataset was manipulated as well,
but the change was restricted to the \emph{labels}.
In this work, the \emph{inputs} are transformed.
This is the first of many works with such a goal.
More specifically, the stated goal of \citet{zemel2013learning} is
to achieve both \emph{group fairness} (in the sense of Demographic Parity)
and \emph{individual fairness} (in the sense of treating similar people similarly),
as \citet{dwork2012fairness} did before.
The learned fair representation is denoted by \(z\).
The condition for fairness is then
\begin{align}
  \label{eq:fair-representation}
  P(s=0|z=z') = P(s=1|z=z') ~.
\end{align}
\citet{zemel2013learning} propose to map the biased inputs \(x\) to \emph{prototypes} in the same space as \(x\).
The probability for being assigned to a particular prototype
must be the same for inputs with \(s=0\) and for \(s=1\).
Based on a given distance function (or similarity measure),
inputs are more likely to be assigned to prototypes that are close-by.
The classification task is then based entirely on the prototypes (the fair representation \(z\)).
In the paper, a linear model is used to map the prototypes to the outcomes \(y\).
Aside from the distance function, the whole model has therefore only two kinds of parameters:
the locations of the prototypes and the weights in the linear model.
These parameters are all optimized together.
The objective function consists of three terms:
the first enforcing Demographic Parity via the prototype locations,
the second ensuring that the prototypes are close to the inputs
and the third targeting a good accuracy by adapting the weights of the classifier.

@@

While the algorithms by \citet{feldman2015certifying} are manually constructed and explicit,
\citet{louizos2016variational} use an approach
that falls more in the area of end-to-end learning where fewer hand-crafted algorithms are used.
The method is based on deep variational autoencoders (VAE).
There is an encoder and a decoder that are both modelled as deep neural networks.
The encoder produces the distribution of the latent (fair) representation \(z\)
from the original features \(x\) and the sensitive attribute \(s\).
The decoder recovers the distribution of \(x\) from \(z\) and \(s\).
By choosing a factorized prior \(P(s)P(z)\), a separation between \(s\) and \(z\) is encouraged.

This method can be improved further by taking into account the labels
when constructing the fair representation.
If this is not done, \(z\) loses the ranking information from \(x\) (see \citet{feldman2015certifying} above).
To this end, a second latent variable is introduced, \(\tilde{z}\),
which encodes the variation in \(z\) that is not explained by the class labels \(y\).
\(z\) is then determined by \(y\) and \(\tilde{z}\),
and \(x\) is determined by \(z\) and \(s\) as before.
\(\tilde{z}\) and \(s\) have independent priors.
This structure ensures that \(y\) can be predicted from \(z\).
However, this introduces a new problem:
if \(y\) is correlated with \(s\), then \(z\) will be as well.
To overcome this, an additional penalty term is introduced that forces \(P(z|s=0)\)
and \(P(z|s=1)\) to be as close as possible.
This is realized with a measure of distance between distributions called \acf{MMD}.

@@

\section{Other fairness criteria}%
\label{improved-definitions-of-fairness}
The early literature on fairness in machine learning used only \ac{DP}
and si\-mi\-la\-ri\-ty-based fairness to judge the fairness of a classifier.
\citet{kleinberg2016inherent} formalized three different fairness conditions.
The authors consider a scenario where the data points \(x\) are sorted into bins \(b\)
and each bin is associated with a prediction score \(f = P(\hat{y}=1|b)\)
where \(\hat{y}\) is the predicted class label and \(x\) the (non-sensitive) features.
\begin{enumerate}
\item
  Calibration within groups:
  If the prediction score for a given bin is \(f\),
  then when considering all the data points with group \(s\) in the bin,
  a fraction of \(f\) of those should have the class label \(y=1\).
  In other words, the prediction score \(f\) is well calibrated with respect to group \(s\).
  This should be the case for all groups.
\item
  Balance for the negative class:
  The true negative rate should be the same for both groups: \(P(\hat{y}=0|y=0,s=0) = P(\hat{y}=0|y=0,s=1)\).
\item
  Balance for the positive class:
  The true positive rate should be the same for both groups: \(P(\hat{y}=1|y=1,s=0) = P(\hat{y}=1|y=1,s=1)\).
\end{enumerate}
These are formalizations of things that have been proposed as definitions of fairness.
The first criterion essentially ensures that the predictor works correctly for both groups.
This prevents a classifier from predicting one group correctly
but always returning a negative answer for the other group.
The second and third put emphasis on negative and positive class labels respectively.
In criterion 2, we allow the classifier to mis-classify those data points with \(y=0\),
but we want to make sure that the misclassification rate is the same for both groups.
In other words, members of the different groups have the same chance
to get a correct classification if they should receive a negative classification.
The same holds for criterion 3 and positive classifications (\(y=1\)).

@@

\citet{hardt2016equality} expand on criterion 3 in \citet{kleinberg2016inherent}
and develop a method to enforce it via post-processing.
Furthermore, an alternative criterion is introduced that is the combination of criterion 2 and 3.
This is also the work that gave criterion 3 the name \emph{Equality of Opportunity}
and that popularized the name \emph{Demographic Parity}.
The treatment of fairness by the authors is explicitly ``oblivious'',
which here means that only the general statistics are known
about the features \(x\), the sensitive attributes and the class labels \(y\).
In particular, there is not enough information available to develop a similarity measure
which could be used to target \emph{individual fairness}.

\citet{hardt2016equality} define the criteria via statistical independence:
\emph{Equalized Odds} refers to the case where \(\hat{y}\) and \(s\) are independent conditional on \(y\).
This is equivalent to the true positive rate and the false positive rate being the same for all groups.
\emph{Equality of Opportunity} is, as mentioned above,
the case where just the true positive rate is the same for all groups.
This means that \(\hat{y}\) and \(s\) are independent conditional on \(y=1\).
One main advantage that these definitions have over Demographic Parity is that a perfect predictor satisfies them.
Demographic Parity is at odds with achieving perfect accuracy.
Furthermore, for Demographic Parity a classifier might be forced to give a positive prediction to a lot of individuals
where it is really not appropriate.

In order to construct a fair classifier out of a \emph{binary predictor}
(giving only binary predictions) via post-processing,
the output is randomized in such a way as to remove the bias.
If the overall probability for a positive predictions of the unfair classifier is given by \(P(\hat{y}=1|s)\),
then the randomized probability for the fair positive predictions \(\tilde{y} = 1\) is given by:
\begin{align}
  \label{eq:hardt}
  P(\tilde{y}=1| x, s) = \sum\limits_{y' \in \{0, 1\}} P(\tilde{y} = 1| \hat{y}=y', s) 
  P(\hat{y}=y'| x, s)
\end{align}
There are 4 free parameters \(P(\tilde{y} = 1| \hat{y}=y', s=s')\) with \(y' \in \{0, 1\}\) and \(s' \in \{0, 1\}\).
As an example, if the unfair predictor predicts \(\hat{y} =0\) for an input with \(s=0\),
then the fair predictor predicts \(\tilde{y} =1\) with probability \(P(\tilde{y} = 1| \hat{y}=0, s=0)\)
which might be non-zero.
In addition to the fairness condition, we also want to enforce accuracy.
To that end, we introduce a loss function \(\ell (\tilde{y}, y)\)
that quantifies the cost of predicting the wrong class label.
The final optimization problem for the 4 free parameters is then to minimize \(\ell\)
under the constraint of Equality of Opportunity or Equalized Odds
and the constraint that the parameters must be valid probabilities.

For a predictor that outputs a \emph{score function},
the post-processing step consists of choosing differing thresholds for \(s=0\) and \(s=1\),
such that the predictions become fair.
If \(f\) is the score, then for a given threshold \(t\) we predict \(\hat{y} = 1\) if \(f > t\).
The two thresholds (one for each group) are found by minimizing \(\ell\) with the fairness constraints, as before.
To satisfy the constraint, it might be necessary to randomize the result.
This is done by using two constraints per group:
if \(f\) is above both, the result is \(\hat{y} = 1\), if it is below both, \(\hat{y} =0\)
and if \(f\) is between the thresholds, then the result is chosen at random.
This method -- as well as the one before --
requires the sensitive attribute for all predictions as it has to choose the right threshold.

The authors prove that with this post-processing,
the Bayes optimal (but unfair) classifier becomes the Bayes optimal fair classifier.
Additionally, they investigate the limitations of the ``oblivious'' approach
where the only available information about the data is the joint distribution over \((y, s, \hat{y})\).
With just this information, it is impossible to uncover the underlying dependency structure that produced the bias.
The authors present two very different scenarios that are indistinguishable with just the joint distribution.
Demographic Parity is also affected by this as it only uses the information from the distribution over \((s, \hat{y})\).
Another shortcoming of Equality of Opportunity and Equalized Odds is
that they carry the assumption that the labels are correct.
Demographic Parity does not rely on this assumption so much.
All-in-all, the fairness definition still has to be chosen by the user based on the situation and the goals.

@@

A conceptually very influential work was \citet{friedler2016possibility}.
While this work does not explicitly present a groundtruth-centric approach to bias,
their concepts of the construct space and the observed space are close
to the idea of the (biased) training distribution and the true underlying distribution.

>>>>

\chapter{Related work}\label{ch:related-work}
%
% =========================
%      old introduction
% =========================
% \section{Introduction}\label{introduction}
% Fairness in Machine Learning is a topic that has recently received increasing attention.
% Machine Learning has become very widespread in our society
% which brings some dangers with it.
% Decisions such as credit eligibility and hiring are increasingly done automatically.
% However, the algorithms that are used in these decisions can have strong biases
% that stem from the training data.
% Often, the algorithms are even more biased than the training data.
% \citet{kamishima2012fairness} give a great illustration for this
% (adapted from \citet{calders2010three}).
% The illustration is based on the Adult/Census Income dataset \citep{kohavi1996scaling}
% which contains records of individuals with thirteen attributes, among them age and gender.
% The task is to predict whether or not the individual earns more than \$50K per year.
% In the dataset, 32\% of the men have a positive label (\ie, they earn more than \$50K),
% but of the women only 11\% have a positive label.
% With this ratio of about 3:1, the dataset is already skewed.
% However, when training a \href{https://en.wikipedia.org/wiki/Naive_Bayes_classifier}{Na\"ive Bayes classifier} on the data,
% the ratio in the predictions is even more unfair:
% the amount of women who get a positive prediction is even lower than 11\%.
% If the Na\"ive Bayes classifier does not get access to the gender information,
% the predictions will be a bit fairer,
% but they are still more unfair than the original dataset labels.
%
% This phenomenon of predictions being more unfair than an already skewed dataset
% is due to the generalization that classifiers have to make.
% It is easier to classify all but the most outstanding women as low income
% than to learn a complicated decision boundary.
% Furthermore, removing the gender information does not solve the problem
% because the gender can be inferred from other data.
%
% The following is a review of the literature on Fairness in Machine Learning.
% Several different approaches to and definition of Fairness are presented.
% The focus is on identifying the major trends in the field from the beginning.
\section{Two views of the dataset bias problem}\label{two-views-of-the-dataset-bias-problem}
As hinted to in the introduction,
there is a divide in the relevant literature in how the problem of dataset bias is approached.
On the one hand, there is the ``ground-truth-centric'' view -- expressed through most of the introduction --
that the training set is biased but the deployment setting is not.
By \emph{deployment setting} I am here referring to the real distribution
that will be encountered by the machine learning system once it is in use.

@@

An instance of this setting has been formalised by \citet{blum2020recovering}.
In their formalization, the training set starts out unbiased,
but then, for samples with \(s=0\), class labels \(y\) are flipped from 1 to 0 with probability \(\nu\).
In addition, samples with \(s=0\) and \(y=0\) are dropped with probability \(1-\beta_\mathit{NEG}\)
and those with \(s=0\) and \(y=1\) are dropped with probability \(1-\beta_\mathit{POS}\).
Meanwhile, the test remains unbiased and is the basis for evaluating our models.

@@

\section{Precursors}\label{precursor}
The first published work to address the problem of biases in machine learning
was arguably the work by \citet{pedreshi2008discrimination}
with the main focus being data mining.
The first fundamental conclusion of the authors
is that it is not enough to remove the discriminatory features
(referred to as \emph{sensitive attributes} in other works) from the data.
That is, if the information about, for example, race
is removed from the data,
there is usually enough background information to reconstruct that attribute.
The experiments were done on the ``German credit dataset''
where the task is to classify people as a good or a bad credit risk.
In their context of data mining,
they distinguish between direct and indirect discrimination:
the former explicitly uses discriminatory features in the premise of the rule,
while the latter uses features that are closely associated with discriminatory features%
\footnote{This is similar to the legal view of direct and indirect discrimination
where direct discrimination when people with a specific protected characteristic are treated worse,
whereas indirect discrimination is a policy that does not make direct references to a protected characteristic
but disproportionally affects people with a specific characteristic.}.
% The latter case can be
% detected by checking the other rules in the dataset.

@@

The authors then define a concrete measure of discrimination (where $\hat{y}$ refers to the prediction of a classifier):
\begin{align}
  \label{eq:disc}
  Disc := P(\hat{y}=1|s =1) - P(\hat{y}=1|s = 0)~.
\end{align}
% (The paper uses a different notation.
% The notation here reflects the notation in later papers.)
It is zero when the individuals in both groups have the same chance to get a positive prediction.
Thus, it measures the violation of \ac{DP} (where the rate of positive predictions would the same for all groups).

@@

The second method in \citet{calders2010three}
is based on training two separate models for each sensitive attributes.
That is, the dataset is split in two
where one half contains all examples with \(s=0\), we call this dataset \(D_0\),
and the other half contains all examples with \(s=1\), called \(D_1\).
For prediction, we choose either the model that was trained on \(D_0\) or the one for \(D_1\),
depending on the sensitive attribute of the example.
This method relies on the availability of the sensitive attribute for prediction,
like the previous method.
The two models are separately tweaked to produce overall fair results.
In order to minimise the effect on the accuracy, the models are tweaked the same amount each,
but in opposite directions.
Conceptually, having two different Na\"ive Bayes model depending on the (binary) sensitive attribute
is equivalent to using one Na\"ive Bayes model
in which the sensitive attribute is connected to all other features.
This means that the bias is in all of the features and not just in the class label.
The latter was the assumption of the previously discussed model.
The difference between these assumptions is not explored in much detail in the paper.

The third methods introduces a new latent variable, called \(L\) in the paper.
\(L\) is intended to be an unbiased target class label
that replaces the biased class label \(y\) from the training data.
\(L\) can only be estimated and is unbiased
in the sense that it is statistically independent from sensitive attribute: \(P(L|s=0) = P(L|s=1)\).
The observed class label \(y\) then only depends on the latent label and the sensitive attribute.
\(L\) is found by using expectation maximization
to iteratively search for a value that maximises the likelihood of the dataset.
The search is restricted by enforcing that \(L\) must be equal to \(y\) except
when \(s=0\) and \(y=0\) or when \(s=1\) and \(y=1\)
because these are the two cases where we expect discrimination.
There is other prior knowledge that can be incorporated into the search.
This method seems to have been the most sophisticated at that point in time.

@@

\begin{enumerate}
\item
  Prejudice:
  a statistical dependency between the sensitive attribute
  and the class label or the (non-sensitive) input features.
  If the dependency is with the class label, then the paper calls it \emph{indirect prejudice}.
  If it is with the non-sensitive features, they call it \emph{latent prejudice}.
  In line with \citet{pedreshi2008discrimination},
  \emph{direct prejudice} refers to classifiers that make direct use of the sensitive attribute.
  Indirect prejudice and latent prejudice can be quantified by
  the mutual information between \(y\) and \(s\) or \(x\) and \(s\) respectively.
  \citet{kamishima2012fairness} do not consider latent prejudice to be immediately harmful,
  but it can be a problem with respect to the protection of personal data and compliance with laws.
\item
  Underestimation:
  non-convergence of machine learning algorithm due to limited amount of data.
  This magnitude of this problem can be estimated
  by considering the difference between the actual training sample distribution
  and the distribution that the machine learning model has internalised.
  This is the underlying cause of models that make predictions
  that are even more unfair than the dataset.
\item
  Negative Legacy:
  sampling bias and wrong labels in the dataset.
  In contrast to the problem of \emph{prejudice},
  \emph{negative legacy} might not be detectable by analysing the dataset.
  \emph{Sampling bias}, meaning that certain data points simply are missing,
  and \emph{wrong labels} can only be corrected if other sources of information are available.
\end{enumerate}

We see that previous works nearly exclusively dealt with the first of these issues, \emph{prejudice},
and more precisely \emph{indirect prejudice}.
\citet{kamishima2012fairness} also focus on \emph{indirect prejudice};
essentially because it is the easiest to deal with, apart from \emph{direct prejudice}.
They use \acf{LR} as the basis for their fair classifier.
Their approach to enforcing fairness is qualitatively different from the previous proposals.
Instead of manipulating the dataset or using explicit algorithm to make a classifier fair,
\citet{kamishima2012fairness} treat the fairness constraint like a regulariser.
A term is added to the objective function
that estimates the mutual information between \(y\) and \(s\)
and gets minimised together with the other parts of the objective function.
This strategy leaves most of the work to the optimization that would need to be run anyway.
A factor in front of the regularization term determines how much to value fairness over accuracy.

@@

The proposed idea is then as follows: a classifier is fair
if and only if the predictive distributions for any two data points
are at least as similar as the two points themselves,
according to a given similarity measure for distributions
and a given similarity measure for data points.
They call this condition the \emph{Lipschitz condition}.
The authors propose two practical similarity measures for distributions
that are well-known in the respective literature.
In order to train a fair classifier,
the Lipschitz condition is then used as a constraint for the optimization.
This is similar to how \citet{kamishima2012fairness} used the fairness criterion as a regulariser.
However, \citet{dwork2012fairness} intend to realise it as a strict constraint.
Furthermore, they show that their criterion is compatible with Demographic Parity
but does not guarantee it in general.
When the two demographic groups are similar according to the individual similarity measure,
then the Lipschitz condition will lead to a good approximation of Demographic Parity.
If the demographic groups are very different,
then the Lipschitz criterion is essentially incompatible with Demographic Parity.

@@

\section{Fair representation}\label{sec:fair-representation}
\citet{zemel2013learning} expand on the idea of creating a \emph{fair representation}.
That is, the features of the data points are transformed such that all bias is removed
but still carry the information necessary for predicting the class label.
This is different from \citet{kamiran2009classifying} where the dataset was manipulated as well,
but the change was restricted to the \emph{labels}.
In this work, the \emph{inputs} are transformed.
This is the first of many works with such a goal.
More specifically, the stated goal of \citet{zemel2013learning} is
to achieve both \emph{group fairness} (in the sense of Demographic Parity)
and \emph{individual fairness} (in the sense of treating similar people similarly),
as \citet{dwork2012fairness} did before.
The learned fair representation is denoted by \(z\).
The condition for fairness is then
\begin{align}
  \label{eq:fair-representation}
  P(s=0|z=z') = P(s=1|z=z') ~.
\end{align}
\citet{zemel2013learning} propose to map the biased inputs \(x\) to \emph{prototypes} in the same space as \(x\).
The probability for being assigned to a particular prototype
must be the same for inputs with \(s=0\) and for \(s=1\).
Based on a given distance function (or similarity measure),
inputs are more likely to be assigned to prototypes that are close-by.
The classification task is then based entirely on the prototypes (the fair representation \(z\)).
In the paper, a linear model is used to map the prototypes to the outcomes \(y\).
Aside from the distance function, the whole model has therefore only two kinds of parameters:
the locations of the prototypes and the weights in the linear model.
These parameters are all optimised together.
The objective function consists of three terms:
the first enforcing Demographic Parity via the prototype locations,
the second ensuring that the prototypes are close to the inputs
and the third targeting a good accuracy by adapting the weights of the classifier.

@@

While the algorithms by \citet{feldman2015certifying} are manually constructed and explicit,
\citet{louizos2016variational} use an approach
that falls more in the area of end-to-end learning where fewer hand-crafted algorithms are used.
The method is based on deep variational autoencoders (VAE).
There is an encoder and a decoder that are both modelled as deep neural networks.
The encoder produces the distribution of the latent (fair) representation \(z\)
from the original features \(x\) and the sensitive attribute \(s\).
The decoder recovers the distribution of \(x\) from \(z\) and \(s\).
By choosing a factorised prior \(P(s)P(z)\), a separation between \(s\) and \(z\) is encouraged.

This method can be improved further by taking into account the labels
when constructing the fair representation.
If this is not done, \(z\) loses the ranking information from \(x\) (see \citet{feldman2015certifying} above).
To this end, a second latent variable is introduced, \(\tilde{z}\),
which encodes the variation in \(z\) that is not explained by the class labels \(y\).
\(z\) is then determined by \(y\) and \(\tilde{z}\),
and \(x\) is determined by \(z\) and \(s\) as before.
\(\tilde{z}\) and \(s\) have independent priors.
This structure ensures that \(y\) can be predicted from \(z\).
However, this introduces a new problem:
if \(y\) is correlated with \(s\), then \(z\) will be as well.
To overcome this, an additional penalty term is introduced that forces \(P(z|s=0)\)
and \(P(z|s=1)\) to be as close as possible.
This is realised with a measure of distance between distributions called \acf{MMD}.

@@

\section{Other fairness criteria}%
\label{improved-definitions-of-fairness}
The early literature on fairness in machine learning used only \ac{DP}
and si\-mi\-la\-ri\-ty-based fairness to judge the fairness of a classifier.
\citet{kleinberg2016inherent} formalised three different fairness conditions.
The authors consider a scenario where the data points \(x\) are sorted into bins \(b\)
and each bin is associated with a prediction score \(f = P(\hat{y}=1|b)\)
where \(\hat{y}\) is the predicted class label and \(x\) the (non-sensitive) features.
\begin{enumerate}
\item
  Calibration within groups:
  If the prediction score for a given bin is \(f\),
  then when considering all the data points with group \(s\) in the bin,
  a fraction of \(f\) of those should have the class label \(y=1\).
  In other words, the prediction score \(f\) is well calibrated with respect to group \(s\).
  This should be the case for all groups.
\item
  Balance for the negative class:
  The true negative rate should be the same for both groups: \(P(\hat{y}=0|y=0,s=0) = P(\hat{y}=0|y=0,s=1)\).
\item
  Balance for the positive class:
  The true positive rate should be the same for both groups: \(P(\hat{y}=1|y=1,s=0) = P(\hat{y}=1|y=1,s=1)\).
\end{enumerate}
These are formalizations of things that have been proposed as definitions of fairness.
The first criterion essentially ensures that the predictor works correctly for both groups.
This prevents a classifier from predicting one group correctly
but always returning a negative answer for the other group.
The second and third put emphasis on negative and positive class labels respectively.
In criterion 2, we allow the classifier to mis-classify those data points with \(y=0\),
but we want to make sure that the misclassification rate is the same for both groups.
In other words, members of the different groups have the same chance
to get a correct classification if they should receive a negative classification.
The same holds for criterion 3 and positive classifications (\(y=1\)).

@@

\citet{hardt2016equality} expand on criterion 3 in \citet{kleinberg2016inherent}
and develop a method to enforce it via post-processing.
Furthermore, an alternative criterion is introduced that is the combination of criterion 2 and 3.
This is also the work that gave criterion 3 the name \emph{Equality of Opportunity}
and that popularised the name \emph{Demographic Parity}.
The treatment of fairness by the authors is explicitly ``oblivious'',
which here means that only the general statistics are known
about the features \(x\), the sensitive attributes and the class labels \(y\).
In particular, there is not enough information available to develop a similarity measure
which could be used to target \emph{individual fairness}.

\citet{hardt2016equality} define the criteria via statistical independence:
\emph{Equalised Odds} refers to the case where \(\hat{y}\) and \(s\) are independent conditional on \(y\).
This is equivalent to the true positive rate and the false positive rate being the same for all groups.
\emph{Equality of Opportunity} is, as mentioned above,
the case where just the true positive rate is the same for all groups.
This means that \(\hat{y}\) and \(s\) are independent conditional on \(y=1\).
One main advantage that these definitions have over Demographic Parity is that a perfect predictor satisfies them.
Demographic Parity is at odds with achieving perfect accuracy.
Furthermore, for Demographic Parity a classifier might be forced to give a positive prediction to a lot of individuals
where it is really not appropriate.

In order to construct a fair classifier out of a \emph{binary predictor}
(giving only binary predictions) via post-processing,
the output is randomised in such a way as to remove the bias.
If the overall probability for a positive predictions of the unfair classifier is given by \(P(\hat{y}=1|s)\),
then the randomised probability for the fair positive predictions \(\tilde{y} = 1\) is given by:
\begin{align}
  \label{eq:hardt}
  P(\tilde{y}=1| x, s) = \sum\limits_{y' \in \{0, 1\}} P(\tilde{y} = 1| \hat{y}=y', s) 
  P(\hat{y}=y'| x, s)
\end{align}
There are 4 free parameters \(P(\tilde{y} = 1| \hat{y}=y', s=s')\) with \(y' \in \{0, 1\}\) and \(s' \in \{0, 1\}\).
As an example, if the unfair predictor predicts \(\hat{y} =0\) for an input with \(s=0\),
then the fair predictor predicts \(\tilde{y} =1\) with probability \(P(\tilde{y} = 1| \hat{y}=0, s=0)\)
which might be non-zero.
In addition to the fairness condition, we also want to enforce accuracy.
To that end, we introduce a loss function \(\ell (\tilde{y}, y)\)
that quantifies the cost of predicting the wrong class label.
The final optimization problem for the 4 free parameters is then to minimise \(\ell\)
under the constraint of Equality of Opportunity or Equalised Odds
and the constraint that the parameters must be valid probabilities.

For a predictor that outputs a \emph{score function},
the post-processing step consists of choosing differing thresholds for \(s=0\) and \(s=1\),
such that the predictions become fair.
If \(f\) is the score, then for a given threshold \(t\) we predict \(\hat{y} = 1\) if \(f > t\).
The two thresholds (one for each group) are found by minimizing \(\ell\) with the fairness constraints, as before.
To satisfy the constraint, it might be necessary to randomise the result.
This is done by using two constraints per group:
if \(f\) is above both, the result is \(\hat{y} = 1\), if it is below both, \(\hat{y} =0\)
and if \(f\) is between the thresholds, then the result is chosen at random.
This method -- as well as the one before --
requires the sensitive attribute for all predictions as it has to choose the right threshold.

The authors prove that with this post-processing,
the Bayes optimal (but unfair) classifier becomes the Bayes optimal fair classifier.
Additionally, they investigate the limitations of the ``oblivious'' approach
where the only available information about the data is the joint distribution over \((y, s, \hat{y})\).
With just this information, it is impossible to uncover the underlying dependency structure that produced the bias.
The authors present two very different scenarios that are indistinguishable with just the joint distribution.
Demographic Parity is also affected by this as it only uses the information from the distribution over \((s, \hat{y})\).
Another shortcoming of Equality of Opportunity and Equalised Odds is
that they carry the assumption that the labels are correct.
Demographic Parity does not rely on this assumption so much.
All-in-all, the fairness definition still has to be chosen by the user based on the situation and the goals.

@@

A conceptually very influential work was \citet{friedler2016possibility}.
While this work does not explicitly present a ground-truth-centric approach to bias,
their concepts of the construct space and the observed space are close
to the idea of the (biased) training distribution and the true underlying distribution.
